Astrophysical calibration of gravitational-wave detectors

### 2015-09-25 docs/cosmic_cal.tex
Some minor edits and fixes of some refs

\newcommand{\btheta}{\mathbf{\theta}}

@@

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Sec.~\ref{sec:cosmo} we show the likely accuracy of 
\ac{sGRB} distances obtained from their host galaxy redshift and in Sec.~\ref{sec:discussion} we 
conclude.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabret-Perot Michelson interferometer; where 
an elaborate feedback system is used to sustain a defined measurement in arm length difference 
between the moving mirrors. A comprehensive description of the calibration procedure (in particular 
for the LIGO detectors during their fifth science run) can be found in \cite{2010NIMPA.624..223A}, 
and a brief description is giving in \cite{Vitale:2012} and references therein. For \ac{aLIGO} 
and \ac{AdV}, at the frequency range used in this project, the error in the amplitude calibration 
using the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed new method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}, which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, Type 1a Supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the detector network. However, measurement of the phase evolution of such an
event allows accurate determination of the chirp mass~\footnote{In reality it
is the redshifted chirp mass that is measured.}. Timing information from the
different signal arrival times at each detector in the network allows
reasonably accurate sky position determination.  Finally, amplitude variation
between differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.  A great help to this final
stage for this project is the assumption that the source has an \ac{sGRB}
counterpart.  It follows that we therefore have a well constrained prior belief
on the inclination of the source orbit with respect to the line-of-sight based
on the expected level of beaming in the \ac{sGRB} emission.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosty distance to sources.
This is distinct from Type 1a Supernovae standard candles that only provide
relative luminosty distance measures and require calibration via other methods
as part of the cosmological distance ladder. At first sight the missing
component for standard sirens is a complementary measurement of the source
redshift which would allow each event to inform the distance-redshift
relationship and hence obtain cosmological parameter estimates.  However, it
was not thought possible (until recently) that redshift measurements could be
obtained from \ac{GW} observations. In~\cite{DelPozzo:2011uf} (building on the
original idea proposed in~\cite{1986Natur.323..310S} the idea of identifying
host galaxies within the sky position uncertainty region was developed. Each
galaxy would provide a redshift measure and would contribute to a disparate mix
of possible cosmological parameters. However, by combining multiple (${\sim}20$)
detections, the true set of cosmological parameters emerges with an expected
${\sim}5\%$ uncertainty on the Hubble constant.

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those \acp{GRB} with duration ${<}2$ sec) are
emitted during the merger of
\acp{BNS}~\cite{1989Natur.340..126E},~\cite{1992ApJ...395L..83N}. The emmission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of \ac{BNS} mergers. If the event
exhibits an optical afterglow then the host galaxy can be identified from which
a redshift can be obtained.  The fraction of events with associated redshifts
is ${\sim}1/3$. The range to which the \emph{Swift} \ac{XRT} has detected
\acp{sGRB} is $z{\sim}2$ and the nearest event with associated redshift is at
$z{sim}0.1$~\cite{2015GCN..17278...1C} which is approximately equal to the
horizon range of the advanced \ac{GW} detector network.

@@

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would be identified is 
through the targeted follow-up of an observed \ac{GW} or by post-facto matching of \ac{GW} 
trigger lists with known \ac{sGRB} events.  The likelihood of being able to follow-up \ac{GW} 
events using gamma-ray telescopes with low enough latency to catch a \ac{sGRB} is low.  Compounding 
this issue is the relatively large \ac{GW} sky error-box giving a field-of-view for the \ac{EM} 
observatories to search spanning $\sim 100$'s of square degrees~\cite{grb}.  For the \ac{GW} 
follow-up of \ac{sGRB} scenario the merger time for \ac{BNS} systems will be estimated from the 
\ac{sGRB} to within a few seconds~\cite{grb}.  This makes the follow-up search less computationally
expensive since it is performed over a smaller range of data using potentially fewer numbers of 
waveform templates.  This computational saving enables the use of a more computationally expensive 
multi-detector coherent scheme rather than the cheaper coincidence methods used in the untargeted 
searches.

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the fact that in order 
for such an event to be observed, the \ac{BNS} system must have had its orbital angular momentum 
vector pointing towards (or away) from the detector.  The actual inclination angle of the system, 
defined as the angle between the orbital angular momentum vector and the line of sight, must be $<$
half of the \ac{sGRB} beaming angle.  This prerequisite property limits us to systems that are 
approximately ``face-on'' and therefore biases us to higher \ac{SNR} signals.  However, as 
discussed earlier, the property of beaming severely impacts the probable rate of such joint 
observations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis}

@@

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\scf_{k}(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (in our case Gaussian) strain noise drawn 
from a, $\btheta$ is the set of waveform parameters and $\scf_k$ is the calibration scale factor for 
the $k^{\rm th}$ detector.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf). We use Bayes' theorem for which we 
need to define a likelihood function and prior pdfs on the signal parameters being estimated. In 
this analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to 
sample the posterior distribution. We use a Gaussian likelihood function given by
\begin{widetext}
\begin{equation}
p(\bm{d} | \btheta, \vec{\scf}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm 
det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\scf_{k}\gws_{k,i}(\btheta) - \frac{1}{2}\left(\scf_k^2 \gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
\end{widetext}
where $\bm{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's measured one sided noise \ac{PSD} (and as such 
will contain the effect of the calibration scale factor), and the $i$ indices increment over 
frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters have been subsumed into a vector $\btheta$ and the 
calibration scale factors are within the vector $\vec{\scf}$.

@@

Finally, we require a prior on the calibration scale factors for each detector. We make the 
assumption that in general the calibration applied to the detector data will be correct, so want to 
use a prior that is peaked at unity. We also assume that the probability density that the 
calibration scale factor is either larger or smaller than unity by an equivalent factor, e.g.\ 
$s=10$ or $s=0.1$, is the same. For the scale factor for each detector we we therefore use a 
log-normal distribution as our prior
\begin{equation}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
where we chose $\sigma = 1.07$, which in turn means that having the mode at unity gives $\mu = 
1.15$. With these parameters the prior probability density at 0.1 and 10 is one tenth of that at 
1. Note that this prior does not give a symmetric amount of probability about unity, but given our 
simulation criterion described below in Section~\ref{sec:simulations} we find that the likelihood 
generally overwhelms the prior. As $\scf$ is a scale factor we also have the further constraint 
that it must be positive.

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 900\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations used the 
two \ac{aLIGO} detectors (H1 and L1) and the \ac{AdV} detector (V1). The simulations were all 
performed in the frequency domain and spanned a frequency range from 20\,Hz to 400\,Hz. We assumed
all detectors were operating at their design sensitivity (as given by Figure~1 of 
\cite{2013arXiv1304.0670L}). As such we added noise to each simulation based on the sensitivity 
curve, but scaled with the associated calibration scale factor.

@@

From these posterior distributions we have calculated the minimal 68\% credible region on the 
calibration scale factors for each detector (if these were Gaussian distributions this is 
equivalent to the region either side of the mean bounded by the $1\sigma$ intervals). For all the 
signals at each distance increment we have produced the distribution of the fractional half widths 
(i.e.\ $1\sigma$) of these scale factors' confidence intervals compared to the actual value. These 
are shown as boxplots in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} 
and \ac{NSBH} systems respectively. The boxes show the extent from the lower to upper quartile of 
the values, whilst the whiskers extend from the 5$^{\rm th}$ to 95$^{\rm th}$ percentile. The black 
line within each box gives the median value and the star gives the mean value. Also shown on each 
plot as the dashed magenta line is the percentage of signals drawn from the prior distribution that 
fulfil the detection criterion.

@@

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a maximum at 
$\sim 25\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $\sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still be correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH} systems although the 
higher \ac{SNR} of the signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on 
the calibration scale factors for H1 and L1 out to greater distances. The distributions generally 
appear slightly broader than those for the \ac{BNS} systems. This may be entirely explicable 
through two factors. The first factor is from the fact that including the spin parameter leads to 
strong correlations between the chirp mass, mass ratio and spin. These strong correlations make 
convergence of the \ac{MCMC} take longer and means we have fewer independent samples to estimate 
the posterior distributions with. This leads to a larger statistical fluctuation on the results. 
The second factor is that there appears to be a modest effect due to the population of sources that 
are detectable. For distances at which the \ac{BNS} and \ac{NSBH} systems would give the same 
percentage of observable sources there is a slight increase in the mean and standard distribution 
of \acp{SNR} of the \ac{NSBH} systems over the \ac{BNS} systems.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Distance estimates from \acp{sGRB}\label{sec:cosmo}}

~\cm{Describe the cosmological analysis used to obtain the distance from the
GRB}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Multiple events without counterpart\label{sec:multiple}}

\cm{A discussion of the issues regarding multiple events
without EM counterparts.}

@@

We have attempted to see how well we can assess the detector calibration for \ac{aLIGO} and 
\ac{AdV} using astrophysical sources. To do this we require that a \ac{sGRB}, for which it has 
been possible to measure the distance, is observed in coincidence with a \ac{CBC} signal in the 
\ac{GW} detectors. This enables us to assume a known distance to the source, and also limit the 
inclination of the source, which in turn allows us to test the consistency of the detector 
calibration. We do this by including an unknown scale factor on the true signal and noise, which we 
estimate given the data. We find that for detectable \ac{BNS} sources the uncertainty of on the 
calibration scale factor could on average be determined to $\lesssim 10\%$ of its true value for the 
\ac{aLIGO} and \ac{AdV} detectors out to 100\,Mpc. This is comparable to the hoped for level of 
calibration accuracy of the detectors. For sources at the standard single detector \ac{BNS} horizon 
distance of $\sim 450$\,Mpc the scale factor could on average be determined to within $\lesssim 
20\%$ of its true value for the \ac{aLIGO} detectors. Similar results were found for \ac{NSBH} 
sources, although sources could be observed out to higher distances.

>>>>

\newcommand{\btheta}{\bm{\theta}}

@@

% What do we do in this paper
In this paper we investigate the feasibility of this approach for single coincident 
\ac{GW}--\ac{sGRB} events and establish the validation power of such a calibration technique.  In 
Sec.~\ref{sec:calibration} we briefly summarize the existing experimental technique for \ac{GW} 
detector calibration and its expected accuracy.  We then review the concept of \ac{GW} standard 
sirens and their proposed coincident \ac{EM} signatures, the \acp{sGRB} in
Secs.~\ref{sec:sirens} and~\ref{sec:GRB}. In Secs.~\ref{sec:analysis} and \ref{sec:results} we 
discuss out analysis method and the results, and in Sec.~\ref{sec:discussion} we 
conclude.

@@

The technique used throughout \ac{GW} research in calibrating the detectors is carried out through 
a complicated system of physical manipulations of the Fabry-P\'{e}rot Michelson interferometer; 
where 
an elaborate feedback system is used to sustain a defined measurement in arm length difference 
between the moving mirrors. A comprehensive description of the calibration procedure (in particular 
for the LIGO detectors during their fifth science run) can be found in \cite{2010NIMPA.624..223A}, 
and a brief description is giving in \cite{Vitale:2012} and references therein. For \ac{aLIGO} 
and \ac{AdV}, at the frequency range used in this project, the error in the amplitude calibration 
using the `hardware injection method' is roughly $10\%$ \cite{Vitale:2012}. This is a benchmark set 
for the error estimation using the proposed method in this project.

% Describe our model for the calibration
The \ac{GW} strain is measured through the differential arm length, $\Delta L$, changes of the 
interferometer via $h(f,t) = \Delta L(f,t) / L$, where $L$ is the full arm length. Calibration is 
required to relate the actual measured interferometer error signal output $e(f)$ to $\Delta L$. 
This relation is known as the length response function, $R(f)$, defined such that
\begin{equation}
\Delta L(f,t) = R(f) e(f, t),
\end{equation}
where where we assume $R$ varies only slowly in time (in comparison to transient signal 
time-scales). Calculation of $R$ requires the measurement of various functions within a control 
feedback loop (see \cite{2010NIMPA.624..223A}), which are subject to uncertainties. In this 
study we will assume an estimate of $R$ is available (although in theory we could take on the role 
of estimating $R$ itself), but that it differs from the truth through some unknown scale factor, 
$\scf$, so that
\begin{equation}\label{eq:scalefactor}
\scf h(f) = h_{\rm m}(f) = \frac{R(f,t) e(f)}{L},
\end{equation}
where $h(f)$ is the true strain and $h_{\rm m}(f)$ is the measured strain. With this definition it
means that if $\scf > 1$ then a signal would appear to have a larger amplitude (e.g.\ be 
closer) than in reality\footnote{However, the \ac{SNR} would be the correct value as the signal and 
the noise will both contain the scale factor.}, whereas if $\scf < 1$ it would appear to have a 
lower amplitude than reality. In this analysis we will simplify the situation by assuming that 
$\scf$ is a constant with respect to frequency, but in future studies that assumption could be 
dropped and $\scf$ could take some functional form, or piece-wise fit, with respect to $f$ (e.g.\ 
in a similar way to the method of \cite{2013PhRvD..88h4044L} used for fitting differences in 
\ac{PSD} estimates).

@@

% Standard siren basics
The idea of \ac{GW} standard sirens is directly analogous to the concept of
standard candles in \ac{EM} astronomy and was first proposed
in~\cite{1986Natur.323..310S}. Unlike the primary \ac{EM} standard candle
event, Type Ia Supernovae, the measured luminosity of a \ac{CBC} in \acp{GW} is
not only a function of distance. It also depends upon the chirp mass (a
function of the component masses) and the binary orientation with respect to
the detector network. However, measurement of the phase evolution of such an
event allows accurate determination of the chirp mass~\footnote{In reality it
is the redshifted chirp mass that is measured.}. Timing information from the
different signal arrival times at each detector in the network allows
reasonably accurate sky position determination.  Finally, amplitude variation
between differently oriented interferometers allows some (weaker) level of
determination of the binary system orientation.  A great help to this final
stage for this project is the assumption that the source has an \ac{sGRB}
counterpart.  It follows that we therefore have a well constrained prior belief
on the inclination of the source orbit with respect to the line-of-sight based
on the expected level of beaming in the \ac{sGRB} emission.

% The classic use of sirens
Standard sirens are clearly a very powerful tool for \ac{GW} cosmology since
they give us a direct measure of the absolute luminosty distance to sources.
This is distinct from Type Ia Supernovae standard candles that only provide
relative luminosty distance measures and require calibration via other methods
as part of the cosmological distance ladder. At first sight the missing
component for standard sirens is a complementary measurement of the source
redshift which would allow each event to inform the distance-redshift
relationship and hence obtain cosmological parameter estimates.  However, it
was not thought possible (until recently) that redshift measurements could be
obtained from \ac{GW} observations. In~\cite{DelPozzo:2011uf} (building on the
original idea proposed in~\cite{1986Natur.323..310S}) the idea of identifying
host galaxies within the sky position uncertainty region was developed. Each
galaxy would provide a redshift measure and would contribute to a disparate mix
of possible cosmological parameters. However, by combining multiple (${\sim}20$)
detections, the true set of cosmological parameters emerges with an expected
${\sim}5\%$ uncertainty on the Hubble constant.

@@

% what are sGRBs 
It is believed that \acp{sGRB} (those \acp{GRB} with duration ${<}2$ sec) are
emitted during the merger of
\ac{BNS} or \ac{NSBH} systems ~\cite{1989Natur.340..126E,1992ApJ...395L..83N}. The emission
from these events is highly beamed along the binary rotation axis and hence
only potentially observable for a fraction of the mergers. If the event
exhibits an optical afterglow then the host galaxy can be identified from which
a redshift can be obtained.  The fraction of events with associated redshifts
is ${\sim}1/3$. The range to which the \emph{Swift} \ac{XRT} has detected
\acp{sGRB} is $z{\sim}2$ and the nearest event with associated redshift is at
$z{\sim}0.1$~\cite{2015GCN..17278...1C} which is approximately equal to the
horizon range of the advanced \ac{GW} detector network.

@@

% observing a coincident event
The most likely scenario in which a coincident \ac{GW}--\ac{sGRB} event would be identified is 
through the targeted follow-up of an observed \ac{GW} or by post-facto matching of \ac{GW} 
trigger lists with known \ac{sGRB} events.  The likelihood of being able to follow-up \ac{GW} 
events using gamma-ray telescopes with low enough latency to catch a \ac{sGRB} is low.  Compounding 
this issue is the relatively large \ac{GW} sky error-box giving a field-of-view for the \ac{EM} 
observatories to search spanning $\sim 100$'s of square degrees~\cite{grb}.  For the \ac{GW} 
follow-up of \ac{sGRB} scenario the merger time for \ac{BNS}/\ac{NSBH} systems will be estimated 
from the \ac{sGRB} to within a few seconds~\cite{grb}.  This makes the follow-up search less 
computationally expensive since it is performed over a smaller range of data using potentially 
fewer numbers of waveform templates.  This computational saving enables the use of a more 
computationally expensive multi-detector coherent scheme rather than the cheaper coincidence 
methods used in the untargeted searches.

% discuss the inclination angle issue
A fortunate consequence of a joint \ac{GW}--\ac{sGRB} observation will be the fact that in order 
for such an event to be observed, the \ac{BNS}/\ac{NSBH} system must have had its orbital angular 
momentum vector pointing towards (or away) from the detector.  The actual inclination angle of the 
system, defined as the angle between the orbital angular momentum vector and the line of sight, 
must be $<$
half of the \ac{sGRB} beaming angle.  This prerequisite property limits us to systems that are 
approximately ``face-on'' and therefore biases us to higher \ac{SNR} signals.  However, as 
discussed earlier, the property of beaming severely impacts the probable rate of such joint 
observations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis}\label{sec:analysis}

@@

% Define the likelihood function we will be using
We consider that the measured data in any detector is defined as
%
\begin{equation}
  \tilde{d}_{k}(f)=\scf_{k}(\tilde{n}_{k}(f)+\gws_{k}(f,\btheta))
\end{equation}
%
where $\tilde{n}(f)$ is the Fourier transform of the true (in our case Gaussian) strain noise, 
$\btheta$ is the set of waveform parameters and $\scf_k$ is the calibration scale factor for 
the $k^{\rm th}$ detector, which we want to estimate.

@@

To assess the ability to estimate the unknown calibration scale factors we want to calculate their 
marginal posterior probability distribution functions (pdf). We use Bayes' theorem for which we 
need to define a likelihood function and prior pdfs on the signal parameters being estimated. In 
this analysis we use the Markov chain Monte Carlo code {\tt emcee} \cite{2013PASP..125..306F} to 
sample the posterior distribution. We use a Gaussian likelihood function given by
\begin{widetext}
\begin{equation}
p(\bm{d} | \btheta, \bm{\scf}, \curlH, I) \propto \exp{\left[4\Delta f \sum_{k=1}^{N_{\rm 
det}}
  \sum_{i=i_{f_{\rm low}}}^{i_{f_{\rm high}}}\frac{\Re{\left\{ \tilde{d}^{*}_{k,i} 
\scf_{k}\gws_{k,i}(\btheta) - \frac{1}{2}\left(\scf_k^2 \gws^*_{k,i}(\btheta)\gws_{k,i}(\btheta) + 
\tilde{d}^*_{k,i}\tilde{d}_{k,i}\right) \right\}}}{S_{k,i}}\right]}
\end{equation}
\end{widetext}
where $\bm{d}$ is an array containing the data for all detectors, $\Delta f$ is the frequency 
resolution, $S_k$ is the $k^{\rm th}$ detector's measured one sided noise \ac{PSD} (and as such 
will contain the effect of the calibration scale factor), and the $i$ indices increment over 
frequency with $i_{f_{\rm low}}$ and $i_{f_{\rm high}}$ corresponding to the lower 
and upper range in frequencies used for our analysis, which were 20\,Hz to 400\,Hz respectively. 
This frequency range was chosen as the vast majority of the \ac{SNR} for the inspiralling signals 
we use can be found within this range, so going to lower or higher frequencies provides very little 
additional information. The source parameters have been subsumed into a vector $\btheta$ and the 
calibration scale factors are within the vector $\bm{\scf}$.

@@

Finally, we require a prior on the calibration scale factors for each detector. We make the 
assumption that in general the calibration applied to the detector data will be correct, so want to 
use a prior that is peaked at unity. We also assume that the probability density that the 
calibration scale factor is either larger or smaller than unity by an equivalent factor, e.g.\ 
$\scf=10$ or $\scf=0.1$, is the same. For the scale factor for each detector we we therefore use a 
log-normal distribution as our prior
\begin{equation}
 p(\scf|I) = \frac{1}{\scf\sigma\sqrt{2\pi}}\exp{\left( -\frac{(\ln{\scf} - \mu)^2}{2\sigma^2} 
\right)},
\end{equation}
where we chose $\sigma = 1.07$, which in turn means that having the mode at unity gives $\mu = 
1.15$. With these parameters the prior probability density at 0.1 and 10 is one tenth of that at 
1. Note that this prior does not give a symmetric amount of probability about unity, but given our 
simulation criterion described below in Section~\ref{sec:simulations} we find that the likelihood 
generally overwhelms the prior. As $\scf$ is a scale factor we also have the further constraint 
that it must be positive.

@@

To estimate how well we can constrain the calibration scaling for each detector in an advanced 
detector network we have performed simulations with injected \ac{BNS} and \ac{NSBH} signals at a 
range of distances (50\,Mpc to 500\,Mpc with 50\,Mpc increments for the \ac{BNS} signals and 
100\,Mpc to 900\,Mpc with 100\,Mpc increments for the \ac{NSBH} signals). The simulations used the 
two \ac{aLIGO} detectors (H1 and L1) and the \ac{AdV} detector (V1). The simulations were all 
performed in the frequency domain and spanned a frequency range from 20\,Hz to 400\,Hz. We assumed
all detectors were operating at their design sensitivity (as given by Figure~1 of 
\cite{2013arXiv1304.0670L}). As such we added coloured Gaussian noise to each simulation based on 
the sensitivity curve, but scaled with the associated calibration scale factor.

@@

From these posterior distributions we have calculated the minimal 68\% credible region for the 
calibration scale factors for each detector (if these were Gaussian distributions this is 
equivalent to the region either side of the mean bounded by the $1\sigma$ intervals). For all the 
signals at each distance increment we have produced the distribution of the fractional half widths 
(i.e.\ $1\sigma$) of these scale factors' confidence intervals compared to the true value. These 
are shown as boxplots in figures~\ref{fig:bnsresults} and \ref{fig:nsbhresults} for the \ac{BNS} 
and \ac{NSBH} systems respectively. The boxes show the extent from the lower to upper quartile of 
the values, whilst the whiskers extend from the 5$^{\rm th}$ to 95$^{\rm th}$ percentile. The black 
line within each box gives the median value and the star gives the mean value. Also shown on each 
plot as the dashed magenta line is the percentage of signals drawn from the prior distribution that 
fulfil the detection criterion.

@@

In figure~\ref{fig:bnsresults} we see that on average the scale factor can be recovered to 
equivalent precision in both \ac{aLIGO} detectors, as would be expected, with uncertainties 
generally within the 10\% range for sources at 100\,Mpc. This is comparable to previous estimates 
of the calibration error in the initial LIGO detectors. An interesting feature is that for 
distances $\gtrsim 250$\,Mpc the upper extent of the uncertainty for H1 and L1 hits a maximum at 
$\sim 25\%$, whilst the width of boxes narrows. This is due to our ``detectability'' criteria, 
in that for all distances we will only be seeing those sources with a high enough \ac{SNR} that we 
would consider them detectable. There will therefore be fewer sources with \ac{SNR} higher than 
this criteria at large distances, giving us a narrower range, and we automatically exclude those 
with \ac{SNR} that are too small thus truncating our uncertainty distribution at the upper end. 
However, this does show that on average for sources that are detectable out to 450\,Mpc we would be 
able to constrain the calibration scale factor uncertainty for the \ac{aLIGO} detectors to $\sim 
20\%$. As the detection criteria will generally come from the \ac{SNR} being high enough within the 
two \ac{aLIGO} instruments this means that the SNR in \ac{AdV} can be small and thus the ability to 
constrain the scale factor for it becomes poor (although it still provides information that the 
calibration is not grossly inaccurate). We also see that true uncertainties achievable for 
recovering the calibration scale factors are indeed very similar to the simple assumption that they 
would be given by $\sim 1/{\rm SNR}$, i.e.\ at the upper end of the distribution the \ac{SNR} in 
the two \ac{aLIGO} detectors will be $\sim 5.5$, which would be expected to produce uncertainties 
of $\lesssim 20\%$. Our results are probably slightly poorer than the \ac{SNR} expectation due to 
their still being correlations with $\iota$ even within the constrained prior range, and slight 
correlations between the scale factors for H1 and L1 as seen in figure~\ref{fig:bnspost}.

In figure~\ref{fig:nsbhresults} we see very similar results for the \ac{NSBH} systems although the 
higher \ac{SNR} of the signals means that we can provide $\sim 20\mbox{--}25\%$ uncertainties on 
the calibration scale factors for H1 and L1 out to greater distances. The distributions generally 
appear slightly broader than those for the \ac{BNS} systems. This may be entirely explicable 
through two factors. The first factor is from the fact that including the spin parameter leads to 
strong correlations between the chirp mass, mass ratio and spin. These strong correlations make 
convergence of the \ac{MCMC} take longer and means we have fewer independent samples to estimate 
the posterior distributions with. This leads to a larger statistical fluctuation on the results. 
The second factor is that there appears to be a modest effect due to the population of sources that 
are detectable. For distances at which the \ac{BNS} and \ac{NSBH} systems would give the same 
percentage of observable sources there is a slight increase in the mean and standard deviation 
of \acp{SNR} of the \ac{NSBH} systems over the \ac{BNS} systems.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Distance estimates from \acp{sGRB}\label{sec:cosmo}}

%~\cm{Describe the cosmological analysis used to obtain the distance from the GRB}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Multiple events without counterpart\label{sec:multiple}}

%\cm{A discussion of the issues regarding multiple events
%without EM counterparts.}

@@

We have attempted to see how well we can assess the detector calibration for \ac{aLIGO} and 
\ac{AdV} using astrophysical sources. To do this we require that a \ac{sGRB}, for which it has 
been possible to measure the distance, is observed in coincidence with a \ac{CBC} signal in the 
\ac{GW} detectors. This enables us to assume a known distance and sky position for the source, and 
also limit the inclination of the source, which in turn allows us to test the consistency of the 
detector calibration. We do this by including an unknown scale factor on the true signal and noise, 
which we estimate given the data. We find that for detectable \ac{BNS} sources the uncertainty 
on the calibration scale factor could on average be determined to $\lesssim 10\%$ of its true value 
for the \ac{aLIGO} and \ac{AdV} detectors out to 100\,Mpc. This is comparable to the hoped for 
level of calibration accuracy of the detectors. For sources at the standard single detector 
\ac{BNS} horizon distance of $\sim 450$\,Mpc the scale factor could on average be determined to 
within $\lesssim 20\%$ of its true value for the \ac{aLIGO} detectors. Similar results were found 
for \ac{NSBH} sources, although sources could be observed out to higher distances.
